---
title: "Annotation Quality Evaluation"
author: "Jan Izquierdo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
    fig_width: 14
    fig_height: 7
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

```{r libs}
library(ggplot2)
library(scales)
library(patchwork)
```

---

# Setup

```{r constants}
# ── Focal species ──────────────────────────────────────────────────────────────
FOCAL <- c(
  "Chlamydomonas reinhardtii",  # Archaeplastida  - control, good data
  "Plasmodium falciparum 3D7",  # SAR/Apicomplexa - IsoQuant fails, merge rescues
  "Leishmania donovani",         # Discoba         - divergent tool performance
  #"Paramecium bursaria",         # SAR/Ciliata     - non-standard genetic code
  "Symbiodinium kawagutii",
  "Babesia duncani"              # SAR/Piroplasmida - geneid_merged > reference
)

# ── Pipeline factor levels (Reference = gold-standard target) ─────────────────
PIPELINE_LEVELS <- c("IsoQuant", "LyRic (regular)", "LyRic (merged)",
                     "Geneid (regular)", "Geneid (merged)", "Reference")

# ── Colour palettes ────────────────────────────────────────────────────────────
SP_COLS <- c(
  "Chlamydomonas reinhardtii" = "#4E79A7",
  "Plasmodium falciparum 3D7" = "#E15759",
  "Leishmania donovani"       = "#59A14F",
  "Symbiodinium kawagutii"       = "#F28E2B",
  "Babesia duncani"           = "#B07AA1"
)

PL_COLS <- c(
  "IsoQuant"         = "#BAB0AC",
  "LyRic (regular)"  = "#F28E2B",
  "LyRic (merged)"   = "#C94A4C",
  "Geneid (regular)" = "#76B7B2",
  "Geneid (merged)"  = "#2C7A4B",
  "Reference"        = "#222222"
)

# ── Column name maps: pipeline label -> overview.tsv column ───────────────────
BUSCO_LIN <- c(
  "IsoQuant"         = "isoquant_lineage_busco",
  "LyRic (regular)"  = "lyric_lineage_completeness_regular",
  "LyRic (merged)"   = "lyric_lineage_completeness_merged",
  "Geneid (regular)" = "geneid_regular_lyric_lineage_completeness",
  "Geneid (merged)"  = "geneid_merged_lyric_lineage_completeness",
  "Reference"        = "reference_prot_busco_custom_lineage"
)

BUSCO_EUK <- c(
  "IsoQuant"         = "isoquant_eukaryote_busco",
  "LyRic (regular)"  = "lyric_eukaryote_completeness_regular",
  "LyRic (merged)"   = "lyric_eukaryote_completeness_merged",
  "Geneid (regular)" = "geneid_regular_lyric_eukaryote_completeness",
  "Geneid (merged)"  = "geneid_merged_lyric_eukaryote_completeness",
  "Reference"        = "reference_prot_busco_eukaryote"
)

GENE_COUNT <- c(
  "IsoQuant"         = "isoquant_gene_count",
  "LyRic (regular)"  = "lyric_gene_count_regular",
  "LyRic (merged)"   = "lyric_gene_count_merged",
  "Geneid (regular)" = "geneid_regular_gene_count",
  "Geneid (merged)"  = "geneid_merged_gene_count",
  "Reference"        = "reference_gene_count"
)

GENE_DENS <- c(
  "IsoQuant"         = "isoquant_gene_density_per_mb",
  "LyRic (regular)"  = "lyric_gene_density_per_mb_regular",
  "LyRic (merged)"   = "lyric_gene_density_per_mb_merged",
  "Geneid (regular)" = "geneid_regular_gene_density_per_mb",
  "Geneid (merged)"  = "geneid_merged_gene_density_per_mb",
  "Reference"        = "reference_gene_density_per_mb"
)

IPG <- c(
  "IsoQuant"         = "isoquant_isoforms_per_gene",
  "LyRic (regular)"  = "lyric_isoforms_per_gene_regular",
  "LyRic (merged)"   = "lyric_isoforms_per_gene_merged",
  "Geneid (regular)" = "geneid_regular_isoforms_per_gene",
  "Geneid (merged)"  = "geneid_merged_isoforms_per_gene",
  "Reference"        = "reference_isoforms_per_gene"
)

COD_FRAC <- c(
  "IsoQuant"        = "isoquant_coding_fraction",
  "LyRic (regular)" = "lyric_coding_fraction_regular",
  "LyRic (merged)"  = "lyric_coding_fraction_merged",
  "Reference"       = "reference_coding_fraction"
)

TX_COUNT <- c(
  "IsoQuant"         = "isoquant_transcript_count",
  "LyRic (regular)"  = "lyric_transcript_count_regular",
  "LyRic (merged)"   = "lyric_transcript_count_merged",
  "Geneid (regular)" = "geneid_regular_transcript_count",
  "Geneid (merged)"  = "geneid_merged_transcript_count",
  "Reference"        = "reference_transcript_count"
)
```

```{r theme}
ptm <- theme_bw(base_size=16) +
  theme(
    text              = element_text(size=16, color="black"),
    axis.text         = element_text(color="black"),
    axis.text.x       = element_text(angle=35, hjust=1),
    plot.title        = element_text(hjust=0.5, face="bold"),
    axis.ticks        = element_line(color="black", linewidth=1.4),
    panel.border      = element_rect(color="black", fill=NA, linewidth=2),
    panel.grid.minor  = element_blank(),
    panel.background  = element_rect(fill="transparent"),
    plot.background   = element_rect(fill="transparent", color=NA),
    strip.background  = element_rect(fill="transparent", color="black"),
    strip.text        = element_text(face="bold", size=13),
    legend.background = element_rect(fill="transparent", colour=NA),
    legend.key        = element_rect(fill="transparent", colour=NA)
  )

save_plot <- function(p, name, w=14, h=7) {
  ggsave(file.path("img", paste0(name, ".png")), p,
         width=w, height=h, dpi=300, bg="transparent")
}
```

```{r helpers}
# Build long-format data from a named column map.
# col_map: named character vector (name = pipeline label, value = column in df).
# Returns one row per species x pipeline with value_col as the metric column.
to_long <- function(dat, col_map, value_col) {
  parts <- lapply(seq_along(col_map), function(i) {
    col   <- col_map[[i]]
    label <- names(col_map)[[i]]
    if (!col %in% colnames(dat)) return(NULL)
    out <- data.frame(species=as.character(dat$species),
                      pipeline=label,
                      stringsAsFactors=FALSE)
    out[[value_col]] <- suppressWarnings(as.numeric(dat[[col]]))
    out
  })
  long <- do.call(rbind, Filter(Negate(is.null), parts))
  long$pipeline <- factor(long$pipeline, levels=PIPELINE_LEVELS)
  long <- long[!is.na(long[[value_col]]), , drop=FALSE]
  rownames(long) <- NULL
  long
}
```

```{r load-data}
df <- read.delim("overview.tsv", sep="\t", check.names=FALSE,
                 stringsAsFactors=FALSE, na.strings=c("NA","","nan","None"))

focal <- df[df$species %in% FOCAL, ]
stopifnot(nrow(focal) == 5)
focal$species <- factor(focal$species, levels=FOCAL)

df$is_focal <- df$species %in% FOCAL
```

---

# 1. Dataset overview

## 1.1 RNA-seq availability by supergroup

```{r rnaseq-sg, fig.height=5}
rna_type <- ifelse(is.na(df$rnaseq_data), "No RNA-seq",
            ifelse(grepl("long", df$rnaseq_data) & grepl("short", df$rnaseq_data),
                   "Short + long-read",
            ifelse(grepl("long", df$rnaseq_data), "Long-read only", "Short-read only")))

avail <- as.data.frame(table(supergroup=df$supergroup, rnaseq_type=rna_type),
                       stringsAsFactors=FALSE)
avail$rnaseq_type <- factor(avail$rnaseq_type,
  levels=c("No RNA-seq","Short-read only","Long-read only","Short + long-read"))

sg_order <- names(sort(tapply(avail$Freq, avail$supergroup, sum), decreasing=TRUE))
avail$supergroup <- factor(avail$supergroup, levels=sg_order)

p_avail <- ggplot(avail, aes(x=supergroup, y=Freq, fill=rnaseq_type)) +
  geom_col(colour="black", linewidth=0.4) +
  scale_fill_manual(
    values=c("No RNA-seq"       ="#d9d9d9",
             "Short-read only"  ="#6baed6",
             "Long-read only"   ="#fd8d3c",
             "Short + long-read"="#31a354"),
    name=NULL) +
  labs(x=NULL, y="Species count",
       title="RNA-seq availability across protist supergroups") +
  ptm +
  theme(legend.position="top")

p_avail
save_plot(p_avail, "01_rnaseq_availability", w=10, h=6)
```

## 1.2 Pipeline coverage

```{r coverage, fig.height=10}
has_iso  <- !is.na(df$isoquant_gene_count)
has_lr   <- !is.na(df$lyric_gene_count_regular)
has_lm   <- !is.na(df$lyric_gene_count_merged)
has_gr   <- !is.na(df$geneid_regular_gene_count)
has_gm   <- !is.na(df$geneid_merged_gene_count)

any_data <- has_iso | has_lr | has_lm | has_gr | has_gm
sub      <- df[any_data, c("species","supergroup")]

cov_long <- rbind(
  data.frame(sub, pipeline="IsoQuant",         has_data=has_iso[any_data]),
  data.frame(sub, pipeline="LyRic (regular)",  has_data=has_lr[any_data]),
  data.frame(sub, pipeline="LyRic (merged)",   has_data=has_lm[any_data]),
  data.frame(sub, pipeline="Geneid (regular)", has_data=has_gr[any_data]),
  data.frame(sub, pipeline="Geneid (merged)",  has_data=has_gm[any_data])
)
cov_long$pipeline <- factor(cov_long$pipeline, levels=PIPELINE_LEVELS)

# Order species by supergroup then total pipelines available
n_pip <- tapply(cov_long$has_data, cov_long$species, sum)
sg_sp <- unique(cov_long[, c("species","supergroup")])
sg_sp <- sg_sp[order(sg_sp$supergroup, -n_pip[sg_sp$species]), ]
cov_long$species <- factor(cov_long$species, levels=rev(sg_sp$species))

p_cov <- ggplot(cov_long, aes(x=pipeline, y=species, fill=has_data)) +
  geom_tile(colour="white", linewidth=0.6) +
  scale_fill_manual(values=c("TRUE"="#2C7A4B","FALSE"="#f0f0f0"),
                    labels=c("TRUE"="Available","FALSE"="Missing"),
                    name=NULL) +
  labs(x=NULL, y=NULL, title="Pipeline output availability per species") +
  ptm +
  theme(axis.text.y=element_text(size=7),
        legend.position="top")

p_cov
save_plot(p_cov, "02_pipeline_coverage", w=8, h=16)
```

---

# 2. Focal species - BUSCO completeness

## 2.1 Lineage BUSCO progression

```{r busco-lin, fig.height=6}
lin_long <- to_long(focal, BUSCO_LIN, "busco")
lin_long$species <- factor(lin_long$species, levels=FOCAL)

p_lin <- ggplot(lin_long, aes(x=pipeline, y=busco, colour=species, group=species)) +
  annotate("rect", xmin=5.5, xmax=6.5, ymin=-Inf, ymax=Inf,
           fill="grey92", alpha=0.8) +
  geom_line(data=lin_long[lin_long$pipeline != "Reference", ],
            linewidth=1.3) +
  geom_point(size=3.5, shape=21, fill="white", stroke=1.5) +
  geom_point(data=lin_long[lin_long$pipeline == "Reference", ],
             size=5, shape=18) +
  geom_text(aes(label=round(busco, 1)), vjust=-1.1, size=3,
            show.legend=FALSE) +
  geom_hline(yintercept=50, linetype="dashed", colour="grey60") +
  scale_colour_manual(values=SP_COLS, name=NULL) +
  scale_y_continuous(limits=c(0,108), breaks=seq(0,100,25)) +
  labs(x=NULL, y="Lineage BUSCO (%)",
       title="Lineage BUSCO completeness across pipeline stages",
       caption="Shaded column = reference annotation (protein mode). Geneid evaluated with LyRic-identified lineage.") +
  ptm +
  theme(legend.position="right")

p_lin
save_plot(p_lin, "03_busco_lineage_progression", w=12, h=6)
```

## 2.2 Eukaryote BUSCO progression

```{r busco-euk, fig.height=6}
euk_long <- to_long(focal, BUSCO_EUK, "busco")
euk_long$species <- factor(euk_long$species, levels=FOCAL)

p_euk <- ggplot(euk_long, aes(x=pipeline, y=busco, colour=species, group=species)) +
  annotate("rect", xmin=5.5, xmax=6.5, ymin=-Inf, ymax=Inf,
           fill="grey92", alpha=0.8) +
  geom_line(data=euk_long[euk_long$pipeline != "Reference", ],
            linewidth=1.3) +
  geom_point(size=3.5, shape=21, fill="white", stroke=1.5) +
  geom_point(data=euk_long[euk_long$pipeline == "Reference", ],
             size=5, shape=18) +
  geom_text(aes(label=round(busco, 1)), vjust=-1.1, size=3,
            show.legend=FALSE) +
  geom_hline(yintercept=50, linetype="dashed", colour="grey60") +
  scale_colour_manual(values=SP_COLS, name=NULL) +
  scale_y_continuous(limits=c(0,108), breaks=seq(0,100,25)) +
  labs(x=NULL, y="Eukaryote BUSCO (%)",
       title="Eukaryote BUSCO completeness across pipeline stages",
       caption="Eukaryote BUSCO is lower for divergent lineages (Discoba, ciliates) - compare same species across pipelines, not across species.") +
  ptm +
  theme(legend.position="right")

p_euk
save_plot(p_euk, "04_busco_eukaryote_progression", w=12, h=6)
```

## 2.3 BUSCO heatmap - focal species

```{r busco-heat, fig.height=5}
lin_long$db  <- "Lineage"
euk_long$db  <- "Eukaryote"
heat_focal   <- rbind(lin_long, euk_long)
heat_focal$db <- factor(heat_focal$db, levels=c("Lineage","Eukaryote"))
heat_focal$species <- factor(heat_focal$species, levels=rev(FOCAL))

p_heat <- ggplot(heat_focal, aes(x=pipeline, y=species, fill=busco)) +
  geom_tile(colour="white", linewidth=0.8) +
  geom_text(aes(label=ifelse(is.na(busco), "-", sprintf("%.1f", busco))),
            size=3.2) +
  facet_wrap(~db) +
  scale_fill_gradientn(
    colours  = c("#d73027","#fee08b","#1a9850"),
    values   = c(0, 0.5, 1),
    limits   = c(0, 100),
    na.value = "grey88",
    name     = "BUSCO (%)") +
  labs(x=NULL, y=NULL, title="BUSCO completeness") +
  ptm +
  theme(panel.grid=element_blank(),
        axis.text.x=element_text(angle=30, hjust=1),
        legend.position="right")

p_heat
save_plot(p_heat, "05_busco_heatmap_focal", w=14, h=5)
```

## 2.4 BUSCO bars per species (lineage)

```{r busco-bars, fig.height=6, results='asis'}
for (sp in FOCAL) {
  d <- lin_long[lin_long$species == sp, ]
  cat(sprintf("\n\n### %s\n\n", sp))

  p <- ggplot(d, aes(x=busco, y=reorder(pipeline, busco), fill=pipeline)) +
    geom_vline(xintercept=50, linetype="dashed", colour="grey60", linewidth=1) +
    geom_col(colour="black", linewidth=0.5, width=0.7) +
    geom_text(aes(label=round(busco, 1)), hjust=-0.2, size=4.5, colour="grey20") +
    scale_x_continuous(limits=c(0,115), breaks=seq(0,100,25)) +
    scale_fill_manual(values=PL_COLS, guide="none") +
    labs(x="Lineage BUSCO (%)", y=NULL, title=sp) +
    ptm +
    theme(axis.text.x=element_text(angle=0, hjust=0.5))

  fname <- gsub("[^A-Za-z0-9]", "_", sp)
  save_plot(p, paste0("06_busco_bars_", fname), w=8, h=4)
  print(p)
}
```

---

# 3. Focal species - Gene model counts

## 3.1 Gene count ratio to reference

Gene count divided by the reference gene count for each species.
The reference bar is always at 1.0x by definition.

```{r gene-ratio, fig.height=6}
gc_long <- to_long(focal, GENE_COUNT, "genes")
gc_long$species <- factor(gc_long$species, levels=FOCAL)

ref_gc      <- setNames(suppressWarnings(as.numeric(focal$reference_gene_count)),
                        as.character(focal$species))
gc_long$ref <- ref_gc[as.character(gc_long$species)]
gc_long$ratio <- gc_long$genes / gc_long$ref

p_ratio <- ggplot(gc_long, aes(x=pipeline, y=ratio, fill=pipeline)) +
  geom_col(colour="black", linewidth=0.4, width=0.7) +
  geom_hline(yintercept=1, linetype="dashed", colour="black", linewidth=1) +
  geom_text(aes(label=sprintf("%.2fx", ratio)), vjust=-0.4, size=3) +
  scale_fill_manual(values=PL_COLS, guide="none") +
  scale_y_continuous(limits=c(0, 3.6), breaks=c(0, 0.5, 1, 1.5, 2, 2.5, 3)) +
  facet_wrap(~species, nrow=1) +
  labs(x=NULL, y="Gene count / reference gene count",
       title="Predicted gene count relative to reference annotation",
       caption="Dashed line = reference (1.0x). Reference bar is always 1.0 by construction.") +
  ptm

p_ratio
save_plot(p_ratio, "07_gene_count_ratio", w=16, h=6)
```

## 3.2 Absolute gene counts

```{r gene-abs, fig.height=6}
p_abs <- ggplot(gc_long, aes(x=pipeline, y=genes, fill=pipeline)) +
  geom_col(colour="black", linewidth=0.4, width=0.7) +
  geom_text(aes(label=label_comma()(genes)), vjust=-0.4, size=2.8) +
  scale_fill_manual(values=PL_COLS, guide="none") +
  scale_y_log10(labels=label_comma()) +
  facet_wrap(~species, nrow=1, scales="free_y") +
  labs(x=NULL, y="Gene count (log₁₀ scale)",
       title="Absolute gene counts per pipeline") +
  ptm

p_abs
save_plot(p_abs, "08_gene_counts_absolute", w=16, h=6)
```

## 3.3 Gene density per Mb

```{r gene-density, fig.height=6}
dens_long <- to_long(focal, GENE_DENS, "density")
dens_long$species <- factor(dens_long$species, levels=FOCAL)

p_dens <- ggplot(dens_long, aes(x=pipeline, y=density, fill=pipeline)) +
  geom_col(colour="black", linewidth=0.4, width=0.7) +
  geom_text(aes(label=round(density, 0)), vjust=-0.4, size=3) +
  scale_fill_manual(values=PL_COLS, guide="none") +
  facet_wrap(~species, nrow=1, scales="free_y") +
  labs(x=NULL, y="Genes per Mb",
       title="Gene density per Mb") +
  ptm

p_dens
save_plot(p_dens, "09_gene_density", w=16, h=6)
```

---

# 4. Focal species - Annotation structure

## 4.1 Isoforms per gene

```{r isoforms, fig.height=6}
ipg_long <- to_long(focal, IPG, "ipg")
ipg_long$species <- factor(ipg_long$species, levels=FOCAL)

p_ipg <- ggplot(ipg_long, aes(x=pipeline, y=ipg, fill=pipeline)) +
  geom_col(colour="black", linewidth=0.4, width=0.7) +
  geom_hline(yintercept=1, linetype="dotted", colour="grey50") +
  geom_text(aes(label=sprintf("%.2f", ipg)), vjust=-0.4, size=3) +
  scale_fill_manual(values=PL_COLS, guide="none") +
  scale_y_continuous(limits=c(0, 6.2)) +
  facet_wrap(~species, nrow=1) +
  labs(x=NULL, y="Transcripts / gene",
       title="Isoforms per gene across stages") +
  ptm

p_ipg
save_plot(p_ipg, "10_isoforms_per_gene", w=16, h=6)
```

## 4.2 Coding fraction

Coding fraction = coding transcripts / all transcripts in the transcriptome.
Available for IsoQuant, LyRic and Reference; Geneid is ab-initio and does not
independently call ORFs.

```{r coding-frac, fig.height=6}
cf_long <- to_long(focal, COD_FRAC, "cf")
cf_long$species <- factor(cf_long$species, levels=FOCAL)

p_cf <- ggplot(cf_long, aes(x=pipeline, y=cf, fill=pipeline)) +
  geom_col(colour="black", linewidth=0.4, width=0.6) +
  geom_text(aes(label=sprintf("%.3f", cf)), vjust=-0.4, size=3.2) +
  scale_fill_manual(values=PL_COLS, guide="none") +
  scale_y_continuous(limits=c(0, 1.12), labels=label_percent()) +
  facet_wrap(~species, nrow=1) +
  labs(x=NULL, y="Coding transcripts / all transcripts",
       title="Coding fraction - proportion of transcripts that are protein-coding",
       caption="High coding fraction = precise but may miss non-coding features.\nIsoQuant ~0.97 vs LyRic ~0.73–0.91 illustrates the precision-recall tradeoff.") +
  ptm +
  theme(axis.text.x=element_text(angle=25, hjust=1))

p_cf
save_plot(p_cf, "11_coding_fraction", w=14, h=6)
```

## 4.3 Transcript counts

```{r tx-counts, fig.height=6}
tx_long <- to_long(focal, TX_COUNT, "tx")
tx_long$species <- factor(tx_long$species, levels=FOCAL)

p_tx <- ggplot(tx_long, aes(x=pipeline, y=tx, fill=pipeline)) +
  geom_col(colour="black", linewidth=0.4, width=0.7) +
  geom_text(aes(label=label_comma()(tx)), vjust=-0.4, size=2.8) +
  scale_fill_manual(values=PL_COLS, guide="none") +
  scale_y_log10(labels=label_comma()) +
  facet_wrap(~species, nrow=1, scales="free_y") +
  labs(x=NULL, y="Transcript count (log₁₀ scale)",
       title="Total transcript count per pipeline") +
  ptm

p_tx
save_plot(p_tx, "12_transcript_counts", w=16, h=6)
```

---

# 5. Special case - Symbiodinium: genetic code effect

Ciliates recode TGA and TAA stop codons as sense amino acids.
Genome-mode BUSCO applies the standard genetic code and therefore misreads those
positions as stop codons, producing an artificially low completeness score.
Protein-mode BUSCO operates on existing protein sequences and is unaffected.
The 4.3x gap between reference genome-mode (17.1%) and protein-mode (73.1%)
BUSCO demonstrates that genome-mode scores are uninformative for ciliates.

```{r paramecium, fig.height=5}
pam <- df[df$species == "Symbiodinium kawagutii", ]

pam_d <- data.frame(
  evaluation = c("Reference\n(genome)", "Reference\n(protein)",
                 "IsoQuant", "LyRic\n(regular)", "LyRic\n(merged)",
                 "Geneid\n(regular)", "Geneid\n(merged)"),
  mode = c("genome","protein","protein","protein","protein","protein","protein"),
  lineage  = c(as.numeric(pam$reference_gen_busco_custom_lineage),
               as.numeric(pam$reference_prot_busco_custom_lineage),
               as.numeric(pam$isoquant_lineage_busco),
               as.numeric(pam$lyric_lineage_completeness_regular),
               as.numeric(pam$lyric_lineage_completeness_merged),
               as.numeric(pam$geneid_regular_lyric_lineage_completeness),
               as.numeric(pam$geneid_merged_lyric_lineage_completeness)),
  eukaryote = c(as.numeric(pam$reference_gen_busco_eukaryote),
                as.numeric(pam$reference_prot_busco_eukaryote),
                as.numeric(pam$isoquant_eukaryote_busco),
                as.numeric(pam$lyric_eukaryote_completeness_regular),
                as.numeric(pam$lyric_eukaryote_completeness_merged),
                as.numeric(pam$geneid_regular_lyric_eukaryote_completeness),
                as.numeric(pam$geneid_merged_lyric_eukaryote_completeness)),
  stringsAsFactors=FALSE
)
pam_d$evaluation <- factor(pam_d$evaluation, levels=pam_d$evaluation)

pam_lin <- data.frame(pam_d[, c("evaluation","mode","lineage")],
                      db="Ciliophora lineage")
pam_euk <- data.frame(pam_d[, c("evaluation","mode","eukaryote")],
                      db="Eukaryote")
names(pam_lin)[3] <- "busco"
names(pam_euk)[3] <- "busco"
pam_long <- rbind(pam_lin, pam_euk)
pam_long$db <- factor(pam_long$db, levels=c("Ciliophora lineage","Eukaryote"))

p_pam <- ggplot(pam_long, aes(x=evaluation, y=busco, fill=mode)) +
  geom_col(colour="black", linewidth=0.4, width=0.7,
           position=position_dodge(width=0.8)) +
  geom_text(aes(label=round(busco, 1)),
            position=position_dodge(width=0.8), vjust=-0.5, size=3.2) +
  geom_vline(xintercept=2.5, linetype="dashed", colour="grey50") +
  scale_fill_manual(values=c("genome"="#E15759","protein"="#4E79A7"),
                    name="BUSCO mode") +
  scale_y_continuous(limits=c(0,110)) +
  facet_wrap(~db) +
  labs(x=NULL, y="BUSCO completeness (%)",
       title=" genetic code effect",
       subtitle="Reference genome BUSCO 17.1% vs protein BUSCO 73.1%: 4.3x gap caused by non-standard genetic code (TGA/TAA -> Gln)",
       caption="Vertical dashed line separates reference annotation from pipeline outputs.") +
  ptm +
  theme(axis.text.x=element_text(angle=25, hjust=1),
        legend.position="top")

p_pam
save_plot(p_pam, "13_paramecium_genetic_code", w=13, h=6)
```

---

# 6. All-species context

## 6.1 BUSCO completeness heatmap - all complete species

```{r busco-all, fig.height=11}
complete_spp <- df$species[
  !is.na(df$isoquant_gene_count) &
  !is.na(df$lyric_gene_count_merged) &
  !is.na(df$geneid_merged_gene_count)]

hall <- df[df$species %in% complete_spp, ]

busco_cols_all <- c(BUSCO_LIN["IsoQuant"], BUSCO_LIN["LyRic (regular)"],
                    BUSCO_LIN["LyRic (merged)"], BUSCO_LIN["Geneid (regular)"],
                    BUSCO_LIN["Geneid (merged)"], BUSCO_LIN["Reference"])

hall_long <- to_long(hall, busco_cols_all, "busco")

# Order species by supergroup then name
sg_ord <- df[df$species %in% complete_spp, c("species","supergroup")]
sg_ord <- sg_ord[order(sg_ord$supergroup, sg_ord$species), ]
hall_long$species <- factor(hall_long$species, levels=rev(sg_ord$species))

p_hall <- ggplot(hall_long, aes(x=pipeline, y=species, fill=busco)) +
  geom_tile(colour="white", linewidth=0.5) +
  geom_text(aes(label=ifelse(is.na(busco),"-", sprintf("%.0f", busco))),
            size=2.6) +
  scale_fill_gradientn(colours=c("#d73027","#fee08b","#1a9850"),
                       values=c(0,0.5,1), limits=c(0,100),
                       na.value="grey90", name="Lineage\nBUSCO (%)") +
  labs(x=NULL, y=NULL,
       title="Lineage BUSCO - complete pipeline output") +
  ptm +
  theme(axis.text.x=element_text(angle=25, hjust=1),
        axis.text.y=element_text(size=8),
        panel.grid=element_blank())

p_hall
save_plot(p_hall, "14_busco_heatmap_all", w=11, h=13)
```

## 6.2 LyRic merge gain

```{r lyric-gain, fig.height=5}
lr_r <- suppressWarnings(as.numeric(df$lyric_lineage_completeness_regular))
lr_m <- suppressWarnings(as.numeric(df$lyric_lineage_completeness_merged))
ok   <- !is.na(lr_r) & !is.na(lr_m)

lg <- data.frame(
  species  = df$species[ok],
  delta    = lr_m[ok] - lr_r[ok],
  is_focal = df$is_focal[ok],
  stringsAsFactors=FALSE
)
lg <- lg[order(lg$delta), ]
lg$species <- factor(lg$species, levels=lg$species)

p_lg <- ggplot(lg, aes(x=species, y=delta, fill=is_focal)) +
  geom_col(colour="black", linewidth=0.3, width=0.8) +
  geom_hline(yintercept=0, colour="black") +
  scale_fill_manual(values=c("TRUE"="#E15759","FALSE"="#4E79A7"), guide="none") +
  scale_y_continuous(breaks=seq(-20,100,20)) +
  labs(x=NULL, y="BUSCO gain (merged - regular, %)",
       title="LyRic: lineage BUSCO gain from merging with reference annotation",
       subtitle=sprintf("n=%d  |  mean=+%.1f%%  |  range: %+.1f%% to %+.1f%%",
                        nrow(lg), mean(lg$delta), min(lg$delta), max(lg$delta)),
       caption="Red bars = focal species. Sorted by gain.") +
  ptm +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank())

p_lg
save_plot(p_lg, "15_lyric_merge_gain", w=13, h=6)
```

## 6.3 Geneid merge gain

```{r geneid-gain, fig.height=5}
gd_r <- suppressWarnings(as.numeric(df$geneid_regular_lyric_lineage_completeness))
gd_m <- suppressWarnings(as.numeric(df$geneid_merged_lyric_lineage_completeness))
ok2  <- !is.na(gd_r) & !is.na(gd_m)

gg <- data.frame(
  species  = df$species[ok2],
  delta    = gd_m[ok2] - gd_r[ok2],
  is_focal = df$is_focal[ok2],
  stringsAsFactors=FALSE
)
gg <- gg[order(gg$delta), ]
gg$species <- factor(gg$species, levels=gg$species)

p_gg <- ggplot(gg, aes(x=species, y=delta, fill=is_focal)) +
  geom_col(colour="black", linewidth=0.3, width=0.8) +
  geom_hline(yintercept=0, colour="black") +
  scale_fill_manual(values=c("TRUE"="#E15759","FALSE"="#2C7A4B"), guide="none") +
  scale_y_continuous(breaks=seq(-20,100,20)) +
  labs(x=NULL, y="BUSCO gain (merged - regular, %)",
       title="Geneid: lineage BUSCO gain from merging with transcriptome annotation",
       subtitle=sprintf("n=%d  |  mean=+%.1f%%  |  range: %+.1f%% to %+.1f%%",
                        nrow(gg), mean(gg$delta), min(gg$delta), max(gg$delta)),
       caption="Red bars = focal species. Sorted by gain.") +
  ptm +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank())

p_gg
save_plot(p_gg, "16_geneid_merge_gain", w=13, h=6)
```

## 6.4 Gene count calibration - distribution by pipeline

Ratio = observed gene count / reference coding transcript count.
Includes a Reference distribution for species that have reference gene count data.

```{r ratio-dist, fig.height=6}
ref_ct <- suppressWarnings(as.numeric(df$reference_coding_transcripts))

make_ratio_col <- function(col_name) {
  vals <- suppressWarnings(as.numeric(df[[col_name]]))
  ok <- !is.na(vals) & !is.na(ref_ct) & ref_ct > 0
  data.frame(species=df$species[ok], pipeline=col_name,
             ratio=vals[ok]/ref_ct[ok], is_focal=df$is_focal[ok],
             stringsAsFactors=FALSE)
}

ratio_dist <- rbind(
  make_ratio_col("isoquant_gene_count"),
  make_ratio_col("lyric_gene_count_regular"),
  make_ratio_col("lyric_gene_count_merged"),
  make_ratio_col("geneid_regular_gene_count"),
  make_ratio_col("geneid_merged_gene_count"),
  make_ratio_col("reference_gene_count")
)

pl_labels <- c(
  "isoquant_gene_count"          = "IsoQuant",
  "lyric_gene_count_regular"     = "LyRic (regular)",
  "lyric_gene_count_merged"      = "LyRic (merged)",
  "geneid_regular_gene_count"    = "Geneid (regular)",
  "geneid_merged_gene_count"     = "Geneid (merged)",
  "reference_gene_count"         = "Reference"
)
ratio_dist$pipeline <- factor(pl_labels[ratio_dist$pipeline],
                               levels=PIPELINE_LEVELS)

p_rdist <- ggplot(ratio_dist, aes(x=pipeline, y=ratio, fill=pipeline)) +
  geom_boxplot(outlier.shape=21, outlier.size=2, colour="black",
               linewidth=0.5, width=0.55) +
  geom_jitter(aes(colour=pipeline), width=0.12, size=1.5, alpha=0.5) +
  geom_hline(yintercept=1, linetype="dashed", colour="black", linewidth=1) +
  scale_fill_manual(values=PL_COLS, guide="none") +
  scale_colour_manual(values=PL_COLS, guide="none") +
  scale_y_log10(breaks=c(0.01,0.05,0.1,0.5,1,2,3,5),
                labels=label_number()) +
  labs(x=NULL, y="Genes predicted / reference coding transcripts (log scale)",
       title="Gene count calibration across all species with reference data",
       caption="Dashed line = 1x (expected). Each point is one species. Reference distribution is reference_gene_count / reference_coding_transcripts.") +
  ptm +
  theme(axis.text.x=element_text(angle=25, hjust=1))

p_rdist
save_plot(p_rdist, "17_gene_ratio_distribution", w=11, h=6)
```

## 6.5 Gene count vs BUSCO - completeness vs calibration scatter

```{r scatter, fig.height=10}
make_scatter_source <- function(gene_col, busco_col, label) {
  genes <- suppressWarnings(as.numeric(df[[gene_col]]))
  busco <- suppressWarnings(as.numeric(df[[busco_col]]))
  ref   <- suppressWarnings(as.numeric(df$reference_coding_transcripts))
  ok    <- !is.na(genes) & !is.na(busco) & !is.na(ref) & ref > 0
  data.frame(
    species  = df$species[ok],
    pipeline = label,
    ratio    = genes[ok] / ref[ok],
    busco    = busco[ok],
    is_focal = df$is_focal[ok],
    stringsAsFactors=FALSE
  )
}

scatter <- rbind(
  make_scatter_source("isoquant_gene_count",       "isoquant_lineage_busco",                      "IsoQuant"),
  make_scatter_source("lyric_gene_count_regular",  "lyric_lineage_completeness_regular",           "LyRic (regular)"),
  make_scatter_source("lyric_gene_count_merged",   "lyric_lineage_completeness_merged",            "LyRic (merged)"),
  make_scatter_source("geneid_regular_gene_count", "geneid_regular_lyric_lineage_completeness",    "Geneid (regular)"),
  make_scatter_source("geneid_merged_gene_count",  "geneid_merged_lyric_lineage_completeness",     "Geneid (merged)"),
  make_scatter_source("reference_gene_count",      "reference_prot_busco_custom_lineage",          "Reference")
)
scatter$pipeline <- factor(scatter$pipeline, levels=PIPELINE_LEVELS)

p_scatter <- ggplot(scatter, aes(x=ratio, y=busco, colour=pipeline)) +
  geom_vline(xintercept=1, linetype="dashed", colour="grey50") +
  geom_point(data=scatter[!scatter$is_focal, ], alpha=0.5, size=2) +
  geom_point(data=scatter[scatter$is_focal, ],  size=4.5, shape=21,
             stroke=1.5, aes(fill=pipeline), colour="black") +
  scale_colour_manual(values=PL_COLS, name="Pipeline") +
  scale_fill_manual(values=PL_COLS, guide="none") +
  scale_x_log10(breaks=c(0.01,0.1,0.5,1,2,5),
                labels=label_number()) +
  scale_y_continuous(limits=c(0,105), breaks=seq(0,100,25)) +
  facet_wrap(~pipeline, nrow=2, ncol=3) +
  labs(x="Gene count / reference coding transcripts (log scale)",
       y="Lineage BUSCO (%)",
       title="Gene count calibration vs annotation completeness",
       caption="Dashed vertical line = 1x expected gene count. Large points = focal species.\nHigh BUSCO + high ratio indicates over-annotation rather than genuine quality improvement.") +
  ptm +
  theme(legend.position="none",
        axis.text.x=element_text(angle=30, hjust=1))

p_scatter
save_plot(p_scatter, "18_gene_busco_scatter", w=12, h=10)
```

## 6.6 Annotation precision vs completeness - coding fraction vs BUSCO

Coding fraction (proportion of transcripts that are protein-coding) measures
precision; BUSCO measures completeness/recall.
High coding fraction + low BUSCO = precise but incomplete.
Low coding fraction + high BUSCO = complete but noisy.

```{r busco-codfrac, fig.height=8}
cf_pips <- names(COD_FRAC)   # IsoQuant, LyRic (regular), LyRic (merged), Reference

blong_cf <- to_long(df, BUSCO_LIN[cf_pips], "busco")
cflong   <- to_long(df, COD_FRAC, "cf")
cf_bi    <- merge(blong_cf, cflong, by=c("species","pipeline"))
cf_bi$is_focal   <- cf_bi$species %in% FOCAL
cf_bi$sp_short   <- sub(" .*", "", cf_bi$species)   # first word for labels

p_cf_busco <- ggplot(cf_bi, aes(x=cf, y=busco, colour=pipeline)) +
  geom_point(data=cf_bi[!cf_bi$is_focal, ], alpha=0.45, size=2.2) +
  geom_point(data=cf_bi[cf_bi$is_focal, ],  size=5, shape=21,
             stroke=1.6, aes(fill=pipeline), colour="black") +
  geom_text(data=cf_bi[cf_bi$is_focal, ],
            aes(label=sp_short), vjust=-0.9, size=2.9, show.legend=FALSE) +
  scale_colour_manual(values=PL_COLS[cf_pips], guide="none") +
  scale_fill_manual(values=PL_COLS[cf_pips], guide="none") +
  scale_x_continuous(labels=label_percent(accuracy=1), breaks=seq(0,1,0.2),
                     limits=c(0,1.05)) +
  scale_y_continuous(limits=c(0,105), breaks=seq(0,100,25)) +
  facet_wrap(~pipeline, nrow=2, ncol=2) +
  labs(x="Coding fraction (protein-coding transcripts / all transcripts)",
       y="Lineage BUSCO (%)",
       title="Precision vs completeness: coding fraction vs lineage BUSCO",
       caption="Large points = focal species (first word of name). Top-right = precise and complete.") +
  ptm +
  theme(axis.text.x=element_text(angle=0, hjust=0.5))

p_cf_busco
save_plot(p_cf_busco, "19_busco_vs_coding_fraction", w=11, h=9)
```

## 6.7 BUSCO change vs gene count change relative to reference - focal species

Both axes are relative to the reference annotation for each species, so the
reference is always at the origin (0 pp, 1x).
Points in the upper half gained BUSCO completeness versus the reference;
points left of the dashed line predicted fewer genes.

```{r busco-gc-delta, fig.height=7}
busco_foc <- to_long(focal, BUSCO_LIN, "busco")
gc_foc    <- to_long(focal, GENE_COUNT, "genes")
both_foc  <- merge(busco_foc, gc_foc, by=c("species","pipeline"))

ref_busco_v <- setNames(
  suppressWarnings(as.numeric(focal$reference_prot_busco_custom_lineage)),
  as.character(focal$species))
ref_gc_v <- setNames(
  suppressWarnings(as.numeric(focal$reference_gene_count)),
  as.character(focal$species))

both_foc$delta_busco <- both_foc$busco  - ref_busco_v[as.character(both_foc$species)]
both_foc$gc_ratio    <- both_foc$genes  / ref_gc_v[as.character(both_foc$species)]
both_foc <- both_foc[both_foc$pipeline != "Reference", ]
both_foc$species  <- factor(both_foc$species, levels=FOCAL)
both_foc$pipeline <- droplevels(both_foc$pipeline)

pl_shapes <- c("IsoQuant"=16, "LyRic (regular)"=15, "LyRic (merged)"=17,
               "Geneid (regular)"=18, "Geneid (merged)"=8)

p_delta <- ggplot(both_foc, aes(x=gc_ratio, y=delta_busco,
                                colour=species, shape=pipeline)) +
  geom_vline(xintercept=1, linetype="dashed", colour="grey55", linewidth=0.9) +
  geom_hline(yintercept=0, linetype="dashed", colour="grey55", linewidth=0.9) +
  geom_point(size=5, stroke=1.2) +
  annotate("text", x=0.18, y= 12, label="Fewer genes\nBUSCO improves",
           size=3, colour="grey40", hjust=0) +
  annotate("text", x=0.18, y=-12, label="Fewer genes\nBUSCO declines",
           size=3, colour="grey40", hjust=0) +
  annotate("text", x=2.8,  y= 12, label="More genes\nBUSCO improves",
           size=3, colour="grey40", hjust=1) +
  annotate("text", x=2.8,  y=-12, label="More genes\nBUSCO declines",
           size=3, colour="grey40", hjust=1) +
  scale_colour_manual(values=SP_COLS, name="Species") +
  scale_shape_manual(values=pl_shapes, name="Pipeline") +
  scale_x_log10(breaks=c(0.2,0.5,1,2,3), labels=label_number(),
                limits=c(0.15,3.2)) +
  labs(x="Gene count / reference gene count (log scale)",
       y="Lineage BUSCO vs reference (percentage points)",
       title="BUSCO change vs gene count change - focal species vs reference",
       caption="Reference is the origin (0 pp, 1x). Each point = one (species, pipeline) pair.") +
  ptm +
  theme(axis.text.x=element_text(angle=0, hjust=0.5),
        legend.position="right")

p_delta
save_plot(p_delta, "20_busco_vs_genecount_delta", w=12, h=7)
```

---

# Session info

```{r session-info}
sessionInfo()
```
