---
title: "ProgressStat"
author: "Jan Izquierdo"
date: "2026-03-26"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
```

# read and load

```{r readData}
# Load only ggplot2 for plotting
library(ggplot2)

# Load data using base R
#data<-read.csv("overview.tsv", stringsAsFactors=TRUE, na.strings="NA", sep="\t")
data<-read.csv("../protist_table/protists.tsv", stringsAsFactors=TRUE, na.strings="", sep="\t")
data<-read.csv("../protist_table/strain_protists.tsv", stringsAsFactors=TRUE, na.strings="", sep="\t")
```

```{r filterLongGenome}
genomes_data<-data[!is.na(data$assembly_accession),]
longread_genomes<-genomes_data[genomes_data$long.read_experiments>0,]

#strains to species
strain_list<-strsplit(trimws(data$species), "\\s+")
#extract valid binomial names
species_vectors <- lapply(strain_list, function(w) {
  if (length(w) < 2) return(NULL) #only larger than 2(no group names(plasmodium))
  
  clean_epithet <- tolower(gsub("\\.$", "", w[2])) #remove .
  if (clean_epithet %in% c("sp", "cf", "aff", "clade")) return(NULL) #remove indicators
  
  return(tolower(paste(w[1], w[2])))
})

#unlist, remove NULLs, count 
allSpecies<-data.frame(Species=unique(unlist(species_vectors)))
```

#stats

```{r}
counts<-c(
  "Protist strains in SRA"=nrow(data),
  "Species"=nrow(allSpecies),
  "Protist Assemblies"=sum(!is.na(data$assembly_accession)),
  "Annotations"=sum(!is.na(data$annotation_accession)),
  #"Strains with RNA-seq"=sum(!is.na(data$rnaseq_data)),
  "Assemblies with
  RNA-seq"=sum(!is.na(data$rnaseq_data) & !is.na(data$assembly_accession)),
  #"Strains with long-read RNA-seq"=sum(data$long.read_experiments > 0, na.rm=TRUE),
  "Assemblies with long-
  read RNA-seq"=sum(data$long.read_experiments > 0 & !is.na(data$assembly_accession), na.rm=TRUE))

plot_data<-data.frame(Category=factor(names(counts)), Count=as.numeric(counts))
```

```{r fig.width=30, fig.height=15}
p<-ggplot(plot_data, aes(x=reorder(Category, -Count), y=Count, fill=Category)) +
  geom_bar(stat="identity", show.legend=FALSE, color="black")+
  geom_text(aes(label=scales::comma(Count)), vjust=-0.5, size=10)+
  scale_fill_brewer(palette="Dark2")+
  labs(x="", y="", title="Protist availability")+
  theme_linedraw()+
  theme(text=element_text(size=33, color="black"),
        axis.text=element_text(color="black"),
        axis.text.x=element_text(angle=45, hjust=1),
        plot.title=element_text(hjust=0.5),
        axis.ticks=element_line(color="black", linewidth=1.5),
        panel.border=element_rect(color="black", fill=NA, linewidth=2),
        #panel.grid.major=element_line(color="black", linewidth=0.8),
        panel.grid.minor=element_line(color="black", linewidth=0.5),
        panel.background=element_rect(fill="transparent"),
        plot.background=element_rect(fill="transparent", color=NA))

p
ggsave("img/protist_availability-assemblies.png", p, width=30, height=20, dpi=400, bg="transparent")
```

```{r}
data$genome_present<-ifelse(!is.na(data$assembly_accession), 1, 0)
data$annotation_present<-ifelse(!is.na(data$annotation_accession), 1, 0)
data$rnaseq_present<-ifelse(!is.na(data$rnaseq_data), 1, 0)
data$longread_present<-ifelse(data$long.read_experiments>0, 1, 0)


#View(data[data$longread_present==1,])
```

#short-long comparison

```{r}
counts<-c(
  "Long-read RNA-seq"=sum(data$long.read_experiments > 0, na.rm=TRUE),
  "Short-read RNA-seq"=sum(data$short.read_experiments > 0, na.rm=TRUE)
  )

plot_data<-data.frame(Category=factor(names(counts)), Count=as.numeric(counts))

p<-ggplot(plot_data, aes(x=reorder(Category, -Count), y=Count, fill=Category)) +
  geom_bar(stat="identity", show.legend=FALSE, color="black")+
  geom_text(aes(label=scales::comma(Count)), vjust=-0.5, size=12)+
  scale_fill_brewer(palette="Dark2")+
  labs(x="", y="")+
  theme_bw()+
  theme(text=element_text(size=35, color="black"),
        axis.text=element_text(color="black", size=40),
        panel.border=element_rect(color="black", fill=NA, linewidth=2),
        axis.ticks=element_line(color="black", linewidth=1.5),
        anel.grid.major=element_line(color="black", linewidth=0.8),
        panel.grid.minor=element_line(color="black", linewidth=0.5),
        panel.background=element_rect(fill='transparent'),
        plot.background=element_rect(fill='transparent', color=NA))
ggsave("img/rnaseq.png", p, width=20, height=20, dpi=700, bg="transparent")
```

#metrics

```{r user-input}
# ── Edit these two variables ──────────────────────────────────────────────────
tsv_path <- "overview.tsv"

chosen_spp <- c(
  "Cyanidioschyzon_merolae_strain_10D",
  "Chlamydomonas_reinhardtii",
  "Babesia_duncani",
  "Emiliania_huxleyi_CCMP1516",
  "Cyanidiococcus_yangmingshanensis",
  "Hydrurus_foetidus",
  "Chlorella_sorokiniana"
)

chosen_spp <- c(
  "Hydrurus_foetidus", "Cyanidiococcus_yangmingshanensis", "Chlorella_sorokiniana", "Emiliania_huxleyi_CCMP1516", "Cyanidioschyzon_merolae_strain_10D"
)
# ─────────────────────────────────────────────────────────────────────────────
```

```{r load-and-filter}
chosen_spp <- gsub("_", " ", chosen_spp)

df_full <- read.delim(tsv_path, sep="\t", stringsAsFactors=FALSE,
                      check.names=FALSE)

df <- df_full[df_full$species %in% chosen_spp, ]

missing <- setdiff(chosen_spp, unique(df$species))
if (length(missing) > 0)
  warning("Species not found and skipped: ", paste(missing, collapse=", "))
if (nrow(df) == 0) stop("None of the requested species were found.")
```

```{r define-columns}
# ── BUSCO scores (0–100 scale) ────────────────────────────────────────────────
busco_cols <- c(
  "reference_gen_busco_custom_lineage",
  "reference_prot_busco_custom_lineage",
  "reference_gen_busco_eukaryote",
  "reference_prot_busco_eukaryote",
  "isoquant_lineage_busco",
  "isoquant_eukaryote_busco",
  "lyric_lineage_busco_regular",
  "lyric_eukaryote_busco_regular",
  "lyric_lineage_busco_merged",
  "lyric_eukaryote_busco_merged"
)

# ── Count columns (integers, log scale) ───────────────────────────────────────
count_cols <- c(
  "reference_gene_count",
  "reference_transcript_count",
  "isoquant_gene_count",
  "isoquant_transcripts_count",
  "lyric_gene_count_regular",
  "lyric_transcript_count_regular",
  "lyric_gene_count_merged",
  "lyric_transcript_count_merged"
)

# Keep only columns that actually exist in the filtered data
busco_cols <- intersect(busco_cols, colnames(df))
count_cols <- intersect(count_cols, colnames(df))
```

```{r parse-factors}
# ── Helper: derive tool / db_type / count_type from raw column names ──────────

add_busco_factors <- function(d) {
  d$tool <- ifelse(grepl("^reference", d$metric), "Reference",
            ifelse(grepl("^isoquant",  d$metric), "IsoQuant",
            ifelse(grepl("^lyric",     d$metric), "Lyric", "Other")))
  d$db_type <- ifelse(grepl("lineage|custom_lineage", d$metric),
                      "Lineage", "Eukaryote")
  d$tool    <- factor(d$tool,    levels=c("Reference", "IsoQuant", "Lyric"))
  d$db_type <- factor(d$db_type, levels=c("Lineage", "Eukaryote"))
  d
}

add_count_factors <- function(d) {
  d$tool <- ifelse(grepl("^reference", d$metric), "Reference",
            ifelse(grepl("^isoquant",  d$metric), "IsoQuant",
            ifelse(grepl("^lyric",     d$metric), "Lyric", "Other")))
  d$count_type <- ifelse(grepl("gene", d$metric), "Gene", "Transcript")
  d$tool       <- factor(d$tool,       levels=c("Reference", "IsoQuant", "Lyric"))
  d$count_type <- factor(d$count_type, levels=c("Gene", "Transcript"))
  d
}

# Alpha values: full opacity=solid category, half=secondary category
ALPHA_BUSCO  <- c("Lineage"=1.0, "Eukaryote"=0.45)
ALPHA_COUNTS <- c("Gene"   =1.0, "Transcript"=0.45)
```

---

## BUSCO scores

One panel per species; metrics run along the y-axis so all scores for a
species are immediately comparable. The dashed line marks the 50 % midpoint.

```{r busco-facet, fig.width=14, fig.height=10}
df_busco <- reshape(df[, c("species", busco_cols)],
                    varying  =busco_cols,
                    v.names  ="score",
                    timevar  ="metric",
                    times    =busco_cols,
                    direction="long")
df_busco <- df_busco[!is.na(df_busco$score), ]
df_busco <- add_busco_factors(df_busco)
df_busco$metric <- gsub("_", " ", df_busco$metric)

p_busco_all <- ggplot(df_busco, aes(x=score, y=reorder(metric, score),
                     fill=tool, alpha=db_type)) +
  geom_vline(xintercept=50, linetype="dashed", colour="grey60", linewidth=1.2) +
  geom_col(colour="black", linewidth=0.5, width=0.7) +
  geom_text(aes(label=round(score, 1)),
            hjust=-0.15, size=6, colour="grey20", alpha=1) +
  scale_x_continuous(limits=c(0, 115), breaks=seq(0, 100, 25)) +
  scale_fill_brewer(palette="Set2", name="Tool") +
  scale_alpha_manual(values=ALPHA_BUSCO, name="DB type",
                     guide =guide_legend(override.aes=list(fill="grey50"))) +
  facet_wrap(~ species, ncol=3, scales="free_y") +
  labs(title="BUSCO Scores by Species", x="Score (%)", y=NULL) +
  theme_bw(base_size=22) +
  theme(
    legend.position   ="top",
    panel.grid.major.y=element_blank(),
    panel.grid.minor  =element_blank(),
    axis.text.y       =element_text(size=18),
    strip.background  =element_rect(fill="transparent", colour="black"),
    strip.text        =element_text(face="bold", size=20),
    panel.border      =element_rect(colour="black", fill=NA, linewidth=2),
    axis.ticks        =element_line(colour="black", linewidth=1.4),
    panel.background  =element_rect(fill="transparent", colour=NA),
    plot.background   =element_rect(fill="transparent", colour=NA),
    legend.background =element_rect(fill="transparent", colour=NA),
    legend.key        =element_rect(fill="transparent", colour=NA)
  )

p_busco_all
ggsave("img/busco_all_species.png", p_busco_all,
       width=28, height=20, dpi=700, bg="transparent")
```

---

## Counts

One panel per species on a log₁₀ scale; all count metrics for a species sit
in the same panel for direct within-species comparison.

```{r counts-facet, fig.width=14, fig.height=10}
df_counts <- reshape(df[, c("species", count_cols)],
                     varying  =count_cols,
                     v.names  ="value",
                     timevar  ="metric",
                     times    =count_cols,
                     direction="long")
df_counts <- df_counts[!is.na(df_counts$value) & df_counts$value > 0, ]
df_counts <- add_count_factors(df_counts)
df_counts$metric <- gsub("_", " ", df_counts$metric)

p_counts_all <- ggplot(df_counts, aes(x=value, y=reorder(metric, value),
                      fill=tool, alpha=count_type)) +
  geom_col(colour="black", linewidth=0.5, width=0.7) +
  geom_text(aes(label=scales::comma(value)),
            hjust=-0.1, size=6, colour="grey20", alpha=1) +
  scale_x_log10(labels=scales::label_comma(),
                expand =expansion(mult=c(0, 0.30))) +
  scale_fill_brewer(palette="Set2", name="Tool") +
  scale_alpha_manual(values=ALPHA_COUNTS, name="Count type",
                     guide =guide_legend(override.aes=list(fill="grey50"))) +
  facet_wrap(~ species, ncol=3, scales="free_y") +
  labs(title="Gene & Transcript Counts by Species",
       x="Count (log\u2081\u2080 scale)", y=NULL) +
  theme_bw(base_size=22) +
  theme(
    legend.position   ="top",
    panel.grid.major.y=element_blank(),
    panel.grid.minor  =element_blank(),
    axis.text.y       =element_text(size=18),
    strip.background  =element_rect(fill="transparent", colour="black"),
    strip.text        =element_text(face="bold", size=20),
    panel.border      =element_rect(colour="black", fill=NA, linewidth=2),
    axis.ticks        =element_line(colour="black", linewidth=1.4),
    panel.background  =element_rect(fill="transparent", colour=NA),
    plot.background   =element_rect(fill="transparent", colour=NA),
    legend.background =element_rect(fill="transparent", colour=NA),
    legend.key        =element_rect(fill="transparent", colour=NA)
  )

p_counts_all
ggsave("img/counts_all_species.png", p_counts_all,
       width=28, height=20, dpi=700, bg="transparent")
```

---

## BUSCO heatmap

Compact overview across all metrics and species at a glance.

```{r busco-heatmap, fig.width=9, fig.height=5}
df_heat <- reshape(df[, c("species", busco_cols)],
                   varying  =busco_cols,
                   v.names  ="score",
                   timevar  ="metric",
                   times    =busco_cols,
                   direction="long")
df_heat$metric <- gsub("_", " ", df_heat$metric)

p_heatmap <- ggplot(df_heat, aes(x=metric, y=species, fill=score)) +
  geom_tile(colour="white", linewidth=1.2) +
  geom_text(aes(label=ifelse(is.na(score), "NA", round(score, 1))),
            size=7, colour="grey20") +
  scale_fill_gradientn(
    colours =c("#d73027", "#fee08b", "#1a9850"),
    values  =c(0, 0.5, 1),
    limits  =c(0, 100),
    na.value="grey90",
    name    ="Score (%)"
  ) +
  labs(title="BUSCO Heatmap", x=NULL, y=NULL) +
  theme_minimal(base_size=26) +
  theme(
    axis.text.x      =element_text(angle=30, hjust=1, size=22),
    axis.text.y      =element_text(size=22),
    panel.grid       =element_blank(),
    panel.background =element_rect(fill="transparent", colour=NA),
    plot.background  =element_rect(fill="transparent", colour=NA),
    legend.background=element_rect(fill="transparent", colour=NA),
    legend.key       =element_rect(fill="transparent", colour=NA)
  )

p_heatmap
ggsave("img/busco_heatmap.png", p_heatmap,
       width=22, height=12, dpi=700, bg="transparent")
```


---

# per-species

One page per species: a BUSCO plot and a counts plot printed separately so
each species can be inspected in isolation.

```{r helper-theme, include=FALSE}
# Shared theme used by both plot types below
spp_theme <- function() {
  theme_bw(base_size=26) +
  theme(
    panel.grid.major.y =element_blank(),
    panel.grid.minor   =element_blank(),
    axis.text.y        =element_text(size=22),
    panel.border       =element_rect(colour="black", fill=NA, linewidth=2),
    axis.ticks         =element_line(colour="black", linewidth=1.4),
    panel.background   =element_rect(fill="transparent", colour=NA),
    plot.background    =element_rect(fill="transparent", colour=NA),
    legend.background  =element_rect(fill="transparent", colour=NA),
    legend.key         =element_rect(fill="transparent", colour=NA)
  )
}

tool_fill <- scale_fill_brewer(palette="Set2")
```

```{r per-species-busco, fig.width=10, fig.height=5, results='asis'}
# ── Prepare long BUSCO data ───────────────────────────────────────────────────
df_busco_all <- reshape(df[, c("species", busco_cols)],
                        varying  =busco_cols,
                        v.names  ="score",
                        timevar  ="metric",
                        times    =busco_cols,
                        direction="long")
df_busco_all <- df_busco_all[!is.na(df_busco_all$score), ]
df_busco_all <- add_busco_factors(df_busco_all)
df_busco_all$metric <- gsub("_", " ", df_busco_all$metric)

# ── One BUSCO plot per species ────────────────────────────────────────────────
for (spp in unique(df$species)) {
  d <- df_busco_all[df_busco_all$species == spp, ]
  if (nrow(d) == 0) next

  cat(sprintf("\n\n### %s\n\n", spp))

  p <- ggplot(d, aes(x=score, y=reorder(metric, score),
                     fill=tool, alpha=db_type)) +
    geom_vline(xintercept=50, linetype="dashed",
               colour="grey60", linewidth=1.2) +
    geom_col(colour="black", linewidth=0.5, width=0.7) +
    geom_text(aes(label=round(score, 1)),
              hjust=-0.15, size=7, colour="grey20", alpha=1) +
    scale_x_continuous(limits=c(0, 115), breaks=seq(0, 100, 25)) +
    tool_fill +
    scale_alpha_manual(values=ALPHA_BUSCO, name="DB type",
                       guide =guide_legend(override.aes=list(fill="grey50"))) +
    labs(title=paste("BUSCO Scores -", spp),
         x="Score (%)", y=NULL, fill="Tool") +
    spp_theme() +
    theme(legend.position="top")

  fname <- gsub(" ", "_", spp)
  ggsave(paste0("img/busco_", fname, ".png"), p,
         width=20, height=20, dpi=700, bg="transparent")
  print(p)
}
```

```{r per-species-counts, fig.width=10, fig.height=5, results='asis'}
# ── Prepare long counts data ──────────────────────────────────────────────────
df_counts_all <- reshape(df[, c("species", count_cols)],
                         varying  =count_cols,
                         v.names  ="value",
                         timevar  ="metric",
                         times    =count_cols,
                         direction="long")
df_counts_all <- df_counts_all[!is.na(df_counts_all$value) & df_counts_all$value > 0, ]
df_counts_all <- add_count_factors(df_counts_all)
df_counts_all$metric <- gsub("_", " ", df_counts_all$metric)

# ── One counts plot per species ───────────────────────────────────────────────
for (spp in unique(df$species)) {
  d <- df_counts_all[df_counts_all$species == spp, ]
  if (nrow(d) == 0) next

  cat(sprintf("\n\n### %s\n\n", spp))

  p <- ggplot(d, aes(x=value, y=reorder(metric, value),
                     fill=tool, alpha=count_type)) +
    geom_col(colour="black", linewidth=0.5, width=0.7) +
    geom_text(aes(label=scales::comma(value)),
              hjust=-0.1, size=7, colour="grey20", alpha=1) +
    scale_x_log10(labels=scales::label_comma(),
                  expand =expansion(mult=c(0, 0.30))) +
    tool_fill +
    scale_alpha_manual(values=ALPHA_COUNTS, name="Count type",
                       guide =guide_legend(override.aes=list(fill="grey50"))) +
    labs(title=paste("Counts -", spp),
         x="Count (log10 scale)", y=NULL, fill="Tool") +
    spp_theme() +
    theme(legend.position="top")

  fname <- gsub(" ", "_", spp)
  ggsave(paste0("img/counts_", fname, ".png"), p,
         width=20, height=20, dpi=700, bg="transparent")
  print(p)
}
```
